---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# herdersmovementpatterns

This repository contains the data and code for our paper:

> Henning Teickner, Christian Knoth, Thomas Bartoschek, Kati Kraehnert, Melinda Vigh, Myagmartseren Purevtseren, Munkhnaran Sugar, Edzer Pebesma, (2020). _Patterns in Mongolian Nomadic Household Movement Derived from GPS Trajectories_. Applied Geography.

### How to cite

Please cite this compendium as:

> Henning Teickner, Christian Knoth, Thomas Bartoschek, Kati Kraehnert, Melinda Vigh, Myagmartseren Purevtseren, Munkhnaran Sugar, Edzer Pebesma, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for 'Patterns in Mongolian Nomadic Household Movement Derived from GPS Trajectories'_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://github.com/henningte/herdersmovementpatterns>.

### How to download or install

You can install this compendium as an R package, herdersmovementpatterns, from GitHub with:

```{r installation, eval = FALSE}
remotes::install_github("henningte\herdersmovementpatterns")
```

You will also have to install the package herdersTA:
```{r installation-herdersTA, eval = FALSE}
remotes::install_github("henningte\herdersTA")
```

### How use

You can reproduce the analyses with:
```{r reproduce, eval=FALSE}
rmarkdown::render("./analysis/paper/paper.Rmd")
```

Moreover, herdersmovementpatterns contains a set of functions to compute movement summary indicators from GPS trajectories preprocessed with herdersTA. See `?herdersmovementpatterns::mp_si` for a brief overview.

### Licenses

**Text and figures :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)

**Code :** See the [DESCRIPTION](DESCRIPTION) file

**Data :** 

1. `herdersmovementpattern::d`: [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse.

2. Shapefile for the Mongolian administrative districts (soums) (data in folder "analysis/data/derived_data"): See https://data.humdata.org/dataset/a9b0a8a6-cb14-448e-b35c-aa5eb51b0557. (Freely available for non-commercial humanitarian use). Source: National Statistical Office of Mongolia.

### Contributions

We welcome contributions from everyone. Before you get started, please see our [contributor guidelines](CONTRIBUTING.md). Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

